\documentclass[twocolumn,english]{article}
\usepackage[latin9]{inputenc}
\usepackage[landscape]{geometry}
\geometry{verbose,tmargin=0.5in,bmargin=0.75in,lmargin=0.5in,rmargin=0.5in}
\setlength{\parskip}{0bp}
\setlength{\parindent}{0pt}

\makeatletter



\usepackage{array}
\usepackage{multirow}
\usepackage{amsbsy}




\providecommand{\tabularnewline}{\\}

\setlength{\columnsep}{0.25in}
\usepackage{xcolor}
\usepackage{textcomp}
\usepackage{listings}
\lstset{
  tabsize=2,
  basicstyle=\small\ttfamily,
}



\usepackage{babel}
\usepackage{listings}
\renewcommand{\lstlistingname}{Listing}

\makeatother

\usepackage{babel}
\begin{document}

\title{Reference Sheet for C220 Software Engineering Design}

\date{Autumn 2017}
\maketitle

\section{Test-Driven Development}

\paragraph{Four Elements of Simple Design}

In order of importance:
\begin{enumerate}
\item Behaves correctly.
\item Minimises duplication.
\item Maximises clarity.
\item Has fewer elements.
\end{enumerate}

\paragraph{TDD Cycle}
\begin{enumerate}
\item Write a failing test (API design).
\item Code to pass test (internals design).
\item Refactor (structural design).
\end{enumerate}

\paragraph{Refactoring}

Process of improving the design of a piece of code, without changing
its functionality.
\begin{itemize}
\item Should be applied \emph{little and often} to continuously improve
design.
\item Only refactor in a \emph{green state}. Tests ensure that behaviour
is preserved.
\item Should be \emph{automated} to be done quickly and reliably.
\item Small transformations are \emph{combined} to achieve larger refactorings.
\end{itemize}

\paragraph{Example Transformations}
\begin{enumerate}
\item \emph{Compose (extract) method}: Break down method into chunks to
make it shorter. Allows us to give a name for a concept and increase
level of abstraction. Try to keep a constant level of abstraction.
\item \emph{Inline variable}: Instead of using a temporary variable, inline
its usages. Reduces number of elements in method.
\item \emph{Extract to common class}: First work to make duplication exactly
the same. Then refactor it, e.g. to another object. Reduces duplication.
\end{enumerate}

\paragraph{Technical Debt}

Features are been added quickly, in an inelegant way. When not fixed
quickly, \emph{technical debt} builds up.
\end{document}
